WFH and broadband speed (title needs rework)

### 2020-11-27 paper/v1/broadband.speed.covid.tex
after conflicts resolved, addition in intro, lit rev, methods and data.

\begin{keywords}
covid; internet; working from home; broadband speed; time series
clusters
\end{keywords}

@@

\begin{itemize}
\tightlist
\item
  \href{https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit\#heading=h.i5om1o8wpcd9}{our
  google doc}
\end{itemize}

During the pandemic, working from home using Information and
Communication Technologies (ICT), whether partially or exclusively, was
transformed from a niche means of accessing work, albeit one that had
been on a slow, upward trend, to a widespread way of life in many
countries \citep{felstead2020homeworking}. \textbf{HANNAH, DO WE REALLY
NEED A REFERENCE IN THE PREVIOUS SENTENCE?} The ability to work from
home or telecommute meant millions retained their jobs and, to a varying
extent, maintained productivity during periods of strict lockdown around
the world. However, this ability has not been evenly distributed
socially or spatially, creating a new type of digital divide. On one
side are those who can work from home and have been able to enjoy both
economic resilience and greater personal safety. On the other side,
previously employed individuals have been forced to accept furlough or
redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection. Using the UK as
a case study, this paper aims to improve our understanding of the
spatial and social dimensions of this new digital divide by assessing
the resilience of broadband speeds in terms of both quality and
reliability of service, and whether this reinforces or redresses prior
digital divisions. To do so, we employ unique volunteered geographic
data on individual broadband speed tests and state-of-the-art
time-series clustering methods, which enable us to create clusters of UK
local authorities with similar temporal signatures of experienced
internet speeds. We then associate these clusters of local authorities
with their socioeconomic and geographic characteristics to explore
whether they are linked with existing divides and the economic geography
of the UK. \textbf{HANNAH, HOW DOES THIS SOUND?} Our analysis enables us
to better understand how telecommuting and technology intersect at a
time of extreme demand, and what lessons this time has for a future
where telecommuting is likely to remain a common means of accessing work
and broadband services, as well as infrastructure, must be fit for
purpose. \textbf{LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE
MORE THE RQ}

The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and
are allowed to be performed independently of location or co-location
with colleagues, including supervisors (\textbf{insert refs}). However,
successful telecommuting also requires that the quality and reliability
of ICT services, particularly home internet connection speeds, enable
the completion of work tasks with a minimum of delay or interruption. In
reality, the performance of broadband speeds with respect to
telecommuters has never been tested before at scale, as working from
home and connecting to colleagues and workplace resources via broadband
has previously been the purview of a small minority of workers. This
papers understands telecommuting as a function of the quality of ICT
services, particularly home internet connections, and whether work tasks
in an occupation can be performed independently of being in a particular
location or co-locating with colleagues

The shift towards telecommuting during various stages of lockdown around
the world has been drastic and, importantly, speculations indicate that
the post-Covid tendency to work from home will be much higher than the
pre-Covid one. A back of the envelope calculation suggests that up to
40\% of the working force could work from home in the UK
\citep{batty2020editorial}. Observational data pointed an even higher
share of people in employment in the UK who worked from home in April
\(2020\) (\(47\)\%), while the same figure only reached \(5\)\% the year
before \citep{ons2020, ons2020lm2019}. Similar figures have been
reported for other countries around the world
\citep{felstead2020homeworking}. For instance, \(37\)\% of the workforce
worked from home in Europe in April \(2020\) with countries like Finland
reaching \(60\)\% \citep{eurofound2020}. In the US almost half of the
working population worked from home during the same period because of
the pandemic \citep{brynjolfsson2020covid}. More broadly, a recent
estimate indicated that \(37\)\% of all jobs in the US can be
permanently performed entirely from home \citep{NBERw26948}.

There is a consensus that opportunities for working from home especially
during the current pandemic are not equally spread across the workforce.
\citet{NBERw26948} indicated that in the US managers, educators, as well
as those working in computer-related occupations, finance, and law can
easily work from home. On the contrary, workers in farming,
construction, and manufacturing do not have such opportunities. Not
surprisingly, occupations with opportunities to telecommute are
associated with higher earnings. This is not the case for the workforce
occupied in less footloose occupations as they tend to be lower-income,
non-white, without a university degree, live in rental accommodation and
lack health insurance \citep{NBERw27085}. Although these figures refer
to the US, similar trends can be observed for other countries. For
example, \(75\)\% of workers with tertiary education worked from home in
Europe during spring \(2020\), while the same share for workers with
secondary and primary education dropped to \(34\)\% and \(14\)\%
respectively. Moreover, employees living in cities, women and younger
employees were have worked from home \citep{eurofound2020}.

None of these changes could have happened in the absence of reliable ICT
infrastructure -- both in terms of software and hardware. But while
software innovations are easily diffused across space and
society\footnote{See for example the huge success of videoconferencing
  apps \citep{marks2020zoom}.}, the same does not apply for ICT hardware
infrastructure such as internet broadband connectivity. The literature
exemplifies digital divides in terms of internet access and its quality.
For instance, \citet{riddlesden2014broadband} highlighted the broadband
divides in the UK, while the systematic review from
\citet{SALEMINK2017360} reinforced our understanding for the
infrastructure quality differences between urban and rural areas.

Our framework to understand telecommuting as a function of occupations
and quality of ICT infrastructure is aligned with current debates on
digital divides. While the, so-called, first level digital divides are
associated with access and quality of internet connectivity, the second
level ones are linked to the necessary skills to effectively utilise ICT
and the internet \citep{blank2014dimensions, van2011internet}.
Importantly, the capacity to telecommute, which to a certain extend is
related to the first and second level digital divides, leads to
differentiated outcomes regarding the economic resilience of people and
places against the current pandemic. And to the extend that the quality
of the internet infrastructure and occupation variation are spatially
dependent and clustered in space, the spatial footprint of telecommuting
is of great interest. In a way similar to the third level digital
divide, which focuses on the differentiated returns of internet use
\citep{stern2009levels, van2014digital, van2015third} places with high
rates of telecommuting during the Covid pandemic illustrate higher
economic resilience against the current pandemic. This is aligned with
the regional economic resilience literature, which underlines the
differentiated capacity of cities and regions to escape or recover from
economic crises \citep{martin2012regional, kitsos2018economic}.

The long-term effects of such drastic changes in telecommuting and
attitudes towards working from home are difficult to predict.
Nevertheless, they span through various aspects of economy and society:
from changes to transportation planning due to altered commuting
patters, to changes in land use and urban planning to accommodate people
who work from home \citep{BUDNITZ2020102713}; and from productivity and
innovation changes, to changes in agglomeration externalities and the
attraction of large cities \citep{econobs} just to name a few. This
paper is positioned to support endeavours in understanding the effects
of increased telecommuting by exposing the spatial and social dimensions
of telecommuting including the resilience of broadband speeds in terms
of both quality and reliability of service, and whether this reinforces
or redresses prior digital divisions. \textbf{HANNAH, HOW DOES THIS
SOUND}

The structure of this paper is as follows. \ldots{}

@@

\hypertarget{broadband-studies-divides-broadband-tech-stuff}{%
\subsection{broadband studies, divides, broadband tech
stuff}\label{broadband-studies-divides-broadband-tech-stuff}}

\begin{itemize}
\tightlist
\item
  PARA about Contention??
\end{itemize}

The infrastructural demands of working from home on internet services
have been minor compared to the demands of leisure users, such that the
broadband performance offered by different Internet Service Providers
(ISPs) has been benchmarked according to download speeds during the
evening `primetime', when video streaming services are at their peak,
rather than during the working day (\textbf{OfCom, 2017}). Yet the
pandemic has fundamentally changed not only how many people work from
home, but also their technical requirements when doing so. Whilst it may
have been common in the past for telecommuters to complete solitary work
tasks at home, but still visit their workplace for meetings
(\textbf{ref?}), during the pandemic the replacement of face-to-face
contact with bandwidth-intensive video conference calling was seen as
essential to suppressing the spread of infection. This type of broadband
use can particularly affect upload speeds \textbf{ref?}. This paper
offers a framework to assess the quality and reliability of upload
speeds by time of day and day of the week across local authorities in
the UK during the time when the population were told to work from home
if at all possible.

\hypertarget{from-telecommuting-to-wfh}{%
\subsection{from telecommuting to
\#WFH}\label{from-telecommuting-to-wfh}}

\begin{itemize}
\tightlist
\item
  Introduce space-time geography
\item
  circumstances where work can be carried out more flexibly in space and
  time
\item
  how relate to digital divides
\item
  capability theory of mobility? suppressed demand\ldots{}
\end{itemize}

In this analysis, the terms `teleworking' and `working from home' are
used interchangeably. An appropriate definition of teleworking is `the
remote provision of labour that would otherwise be carried out within
company premises' (European Commission, 2020b). In practice, during the
COVID crisis, most such work was carried out in the homes of individual
employees rather than any other location. \citep{eurofound2020}

\hypertarget{covid-and-working-from-home-cities-urban-structure}{%
\subsection{covid and working from home, cities, urban
structure}\label{covid-and-working-from-home-cities-urban-structure}}

\begin{itemize}
\tightlist
\item
  economic geography of UK
\end{itemize}

\hypertarget{data-and-descriptive-statistics}{%
\section{Data and descriptive
statistics}\label{data-and-descriptive-statistics}}

The experience of upload speeds is not the same as the maximum speeds
offered by an ISP or possible speeds over a particular type and length
of connection. Therefore, we use volunteered geographic information in
the form of speed-checks run by users to test their experienced
broadband speeds, upload and download. Meanwhile, the quality and
reliability of upload speeds, like ICT more generally, vary in time and
space due to both supply and demand-side influences, and can be measured
in a number of ways. These include: a) mean, experienced connection
speed, b) standard deviation or the amount of fluctuation from the mean,
and c) the variation in speeds at particular times of day when working
from home is more likely to take place. We take account of all three
measurements in order to describe upload speeds as fully as possible. We
start by calculating the mean upload speed for every local government
district in the UK for each hour of each weekday, excluding midnight to
6:00, weekends and bank holidays. We then cluster the local authorities
by these temporal profiles, allowing us to identify patterns and
describe the overall means, standard deviations, and other relevant
statistics for each spatial cluster. Thus we aim to answer the first
part of our research question: How resilient are broadband speeds as
experienced in different parts of the UK during a time of extreme
demand?

The cause of these different experiences of broadband resilience may be
different in different areas, as they may reflect similarities in
patterns of demand or similar quality of infrastructure. Our approach is
also limited by potential endogeneity, as for example, better quality
connections with high mean speeds may enable more working from home, but
greater demand may cause slower speeds, less reliability and greater
variability of speed at different times of day or week. Therefore, we
avoid attributing any cause to our analysis of the experienced level of
quality and reliability of upload speeds. Instead, we run auxilliary
regressions in order to understand how the spatial and temporal patterns
of internet service relate to the economic geography of the UK. We
discuss how the different patterns might support or undermine efforts to
work from home and maintain safe productivity and whether they reinforce
existing spatial and social inequalities. From this analysis, we hope to
provide a greater understanding of how telecommuting and technology
intersect at a time of extreme demand, and what lessons this time has
for a future where telecommuting is likely to remain a common means of
accessing work and broadband services, as well as infrastructure, must
be fit for purpose.

The primary data analysed in this paper was provided by Speedchecker
Ltd, a private company that allows internet users to check their own
broadband upload / download speeds, and stores every speed-check with a
timestamp and geolocation.

Our approach involves aggregating all the speed-checks during the 13
weeks of March to May inclusive for weekdays in 2020 by each hour of the
day and day of the week. As our research aims to identify the geography
of internet service resilience for work purposes, bank holidays and the
hours between midnight and 6am were also excluded. The composite week
time series thus comprise 18 hours multiplied by 5 weekdays or 90 time
points per series. These time series were calculated for each of the 382
Local Authority Districts (LADs) in the UK, standardised, and then a
k-means clustering algorithm was applied, including dynamic time
warping. The LADs were assigned to 10 clusters for upload speeds. The
cluster id for each LAD is then reattached to the speed-check dataframe
to identify the characteristics of each cluster, including number of
LADs, and the descriptives statistics of upload speeds in that cluster,
and the temporal profile by hour of the day and day of the week.

@@

\hypertarget{time-series-clustering}{%
\section{Time series clustering}\label{time-series-clustering}}

Description of the method

@@

The temporal profiles used to cluster the local authorities have been
summarised in graph {[}.{]}, which shows a composite profile of mean
upload speeds per hour per day for each cluster. For upload speeds, 345
of 382 local authorities, or over 62 million people, fall into cluster 6
or cluster 9. Graph {[}.{]} shows that both of these clusters have
relatively similar temporal profiles, which are flatter than the other,
smaller clusters, suggesting better reliability of service. However, the
upload speeds at all times for cluster 9 are substantially higher than
for cluster 6, which is an indication of better quality of service. This
difference is reflected not only in the mean speeds for these clusters
for the whole sample, but also the mean upload speeds in the morning
peak from 9:00-10:59, as well as the evening peak period from
19:00-20:59. In comparison, the time profile in graph {[}.{]} shows
upload speeds in cluster 1 are on average lower at certain times of day
during the study period than any other cluster, whilst the profile for
cluster 3 appears to show speeds fluctuating as much as cluster 1, but
at levels usually higher even than cluster 9.

>>>>

\begin{keywords}
covid; internet; working from home; broadband speed; time-series
clusters
\end{keywords}

@@

During the pandemic, working from home using digital technologies,
whether partially or exclusively, was transformed from a niche means of
accessing work, albeit one that had been on a slow, upward trend, to a
widespread way of life in many countries. The ability to work from home
or telecommute meant millions retained their jobs and, to a varying
extent, maintained productivity during periods of strict lockdown around
the world. However, this ability has not been evenly distributed
socially or spatially, creating a new type of digital divide. On one
side are those who can work from home, supported by digital
technologies, and have thus been able to enjoy both economic resilience
and greater personal safety. On the other side, previously employed
individuals have been forced to accept furlough or redundancy packages
unless they are part of the cadre of essential workers, who are
potentially at high risk of infection. Whilst the basis for this new
digital divide has been viewed as mainly occupational, here we consider
whether the divide is also technological.

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds, may reinforce or redress the spatial
and social dimensions of the digital division exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. \textbf{need to add
urban-rural to analysis} Our analysis enables us to better understand
how the spatial and social distribution of both occupations and online
accessibility intersect to enable or hinder the practice of
telecommuting at a time of extreme demand. We will also consider what
lessons can be learned from this time for a future where telecommuting
is likely to remain a common means of accessing work, and broadband
services and infrastructure must be fit for purpose. \textbf{LET'S LEAVE
IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ}

The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and
are allowed to be performed using digital technologies independently of
location or co-location with colleagues, including supervisors
\citep{allen2015effective, singh2013modeling}. However, successful
telecommuting also requires that the quality and reliability of digital
services, particularly home internet connection speeds, enable the
completion of work tasks with a minimum of delay or interruption. Prior
to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and
connecting to colleagues and workplace resources via the internet was
the purview of a small minority of workers. Instead, leisure use in the
evening, when video streaming services are at their peak, has been used
to benchmark broadband performance and service delivery by different
Internet Service Providers (ISPs), at least in the UK \citep{ofcom2017}.
Yet the shift towards telecommuting during various stages of lockdown
around the world has been drastic and there are speculations that
post-Covid, the tendency to work from home will be much higher than
pre-Covid, raising questions around whether internet services can
accommodate the increased demand. For example, \(47\)\% of people in
employment in the UK worked soley from home in April \(2020\), whilst
the same figure only reached \(5\)\% the year before
\citep{ons2020, ons2020lm2019}. A back of the envelope calculation
suggests that up to 40\% of the working force could work from home on an
ongoing basis \citep{batty2020editorial}. Similar figures have been
reported for other countries \citep{felstead2020homeworking}. For
instance, \(37\)\% of the European workforce worked from home in April
\(2020\) with countries like Finland reaching \(60\)\%
\citep{eurofound2020}. In the US, almost half of the working population
worked from home during the same period because of the pandemic
\citep{brynjolfsson2020covid}, and a recent estimate indicated that
\(37\)\% of all jobs in the US can be permanently performed entirely
from home \citep{NBERw26948}.

None of these changes could have happened in the absence of reliable
information and communication technology (ICT) infrastructure -- both in
terms of software and hardware. But while software innovations are
easily diffused across space and society\footnote{See for example the
  huge success of videoconferencing apps such as Zoom
  \citep{marks2020zoom}.}, the same does not apply for ICT hardware
infrastructure such as internet broadband connectivity. The literature
describes first level digital divides in terms of the availability and
quality of internet connectivity, such as that manifest in different
geographies in the UK
\citep{riddlesden2014broadband, philip2017digital}. Second level digital
divides consider the presence or lack of the necessary skills to
effectively utilise digital technologies and the internet
\citep{blank2014dimensions, van2011internet}. The third level focuses on
the heterogenous returns of internet usage among different socioeconomic
groups and, consequently, how digital technologies can assist in
bridging or further enhancing existing socioeconomic divides.
\citep{stern2009levels, van2014digital, van2015third}. The capability to
telecommute is related to all three levels of digital divides, but more
importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the
current pandemic. We identify this as a new digital divide, one that
fundamentally alters the potential returns of internet use for the user
and wider community, assumes skills or functions that are present in
some occupations but not in others, and relies upon access to high
quality internet services. \textbf{WE NEED TO CONNECT OUR FINDINGS WITH
THIS AND JUSTIFY THAT THIS IS INDEED A NEW DIVIDE} As the quality of
internet infrastructure and services, as well as variation in
occupations are spatially dependent and clustered in space, our approach
offers a framework for understanding which types of places, are more
likely to land on the right side of this new digital divide. By asking
how resilient broadband speeds are as experienced in different parts of
the UK during a time of extreme demand, we interrogate which places
benefit from the greater economic resilience digital technologies can
offer, not only during the pandemic, but also into the future.

The structure of this paper is as follows. First we review the
literature on telecommuting and digital divides to better understand the
origins of the new digital divide revealed by the pandemic and its
impact on the economic resilience of different places. We then describe
our data and methodology. Our results section first offers
classification of how internet services vary across the UK local
authorities and then assesses whether these clusters replicate or
repudiate other socio-economic and geographic patterns of economic
resilience.

\hypertarget{from-telecommuting-to-wfh}{%
\subsection{From telecommuting to
\#WFH}\label{from-telecommuting-to-wfh}}

In this analysis, the terms `telecommuting' and `working from home' are
used interchangeably, as most remote labour during the Covid-19 crisis
was carried out in the homes of individual employees rather than any
other location \citep{eurofound2020}. However, it should be noted that
previous research has explored how telecommuting can occur in other
places, including satellite offices or on public transport
\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
also used a variety of definitions to measure the level of telecommuting
within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or
part-time, and those who use digital technologies to work remotely
full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. For example, studies from the United States, the
Netherlands, and the UK indicate that telecommuters are most likely to
hold professional, managerial, and technical occupations, where the
workforce is better educated and wealthier, and that there is suppressed
demand among women and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.
Opportunities for working from home during the current pandemic have
likewise not been equally spread across the workforce.
\citet{NBERw26948} indicated that in the US, managers, educators, as
well as those working in computer-related occupations, finance, and law
can easily work from home, and that occupations with opportunities to
telecommute are associated with higher earnings. This is not the case
for the workforce occupied in more spatially fixed occupations, from
farming, construction and manufacturing to hospitality and care
services. In the US, these occupations tend to be lower-income,
non-white, without a university degree, live in rental accommodation and
lack health insurance \citep{NBERw27085}. Similar trends can be observed
for other countries. For example, \(75\)\% of workers with tertiary
education worked from home in Europe during spring \(2020\), whilst only
\(34\)\% of workers with secondary education and \(14\)\% of those
primary education did so \citep{eurofound2020}.

\hypertarget{digital-divides-and-economic-resilience}{%
\subsection{Digital divides and economic
resilience}\label{digital-divides-and-economic-resilience}}

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
These returns have been even greater during the Covid-19 crisis, when
those with the ability to telecommute also have the ability to maintain
their employment whilst protecting their health. However, the success of
these arrangements has been dependent upon the first level digital
divide, which is associated with access and quality of internet
connectivity. For example, the systematic review from
\citet{SALEMINK2017360} highlights the infrastructure quality
differences between urban and rural areas in various advanced economies.
Whether this variation in infrastructure quality affects the spatial
footprint of telecommuting has not previously been measured, although
there are indications that those who purchase high speed connections
consume more data of all sorts and use their connections for a variety
of purposes \citep{hauge2011consumer}. Another study identifies a
correlation between access to internet services and a reduction in
household transport spend across 33 countries \citep{bris2017ict}.
Whether the implication is less travel because of increased
telecommuting, or whether internet access enables more efficient travel,
this finding is an indication of the potential household savings better
internet services offer. Thus, the extreme demand during the pandemic
provides a new opportunity to understand how infrastructure
accessibility, quality, and reliability affects telecommuting,
particularly as working from home during the pandemic required high
volumes of bandwidth-intensive video conferencing in order to avoid the
face-to-face contact that could increase the spread of infection. By
first answering questions about internet service resilience, we can also
refine our understanding of how this has contributed to or reduced the
new digital divide, where economic resilience has been dependent upon
the capability to work from home.

The multi-layered digital divides intersect with materials divides and
the economic geography of the UK. Following the regional economic
resilience literature, which underlines the differentiated capacity of
cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}, different places have
different industrial and occupational profiles, and these affect the
aggregated potential capacity of places for telecommuting. Such profiles
are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide
\citep{martin_north_south}. Various studies have illustrated severe
inequalities between the north and the south part of the UK regarding,
for example, skills and human capital, unemployment, productivity and
prosperity \citep{lee2014grim, mccann2020perceptions, dorling2018peak}.
Some scholars have even argued that the UK suffers some of the highest
level of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. Not only all three levels of
digital divides are, to a certain extend, associated with or shaped by
the geography of the UK, but the intersection of the digital and
material divides affects the capacity of places to overcome, at least
partially, the economic effects of the Covid-19 pandemic. Importantly,
this is the first time that digital technologies became an essential
tool for economic resilience for such a great part of the population.

\hypertarget{other-subsection}{%
\subsection{other subsection?}\label{other-subsection}}

\textbf{I THINK WE HAVE ENOUGH OF A LIT REV} - circumstances where work
can be carried out more flexibly in space and time - how relate to
digital divides - capability theory of mobility? - broadband tech stuff?
broadband studies / divides / resilience

@@

\hypertarget{methods-and-data}{%
\section{Methods and data}\label{methods-and-data}}

The starting point of our methodological framework is cluster analysis,
which can be defined within the modern machine learning framework as an
unsupervised learning task, which involves partitioning unlabelled
observations into homogeneous groups known as clusters
\citep{montero2014tsclust}. The key idea is that observations within
clusters tend to be more similar than observations between clusters.
Clustering is particularly useful for exploratory studies as it
identifies structures within the data \citep{aghabozorgi2015time}.
Therefore, cluster analysis is a widely used family of techniques in
geography \citep{gordon1977classification, everitt1974cluster}. For
instance, clustering methods is the basis of \emph{geodemographics},
which aims to create small area indicators or typologies of
neighbourhoods based on various and some times diverse variables
\citep{SINGLETON2009289, harris2005geodemographics}. Clustering
techniques have also been employed to solve regionalisation problems
\citep{niesterowicz2016}.

Common characteristic of these studies is the cross-sectional nature of
the data they employ. Indeed, most clustering problems in geography deal
with observations that are fixed in time. However, for this paper we are
interested in creating clusters of local authorities in the UK with
similar temporal signatures of experienced internet speeds over time.
Hence, we deviate from the established geographical clustering tools and
employ time-series clustering methods.

Time-series clustering methods have been developed in order to deal with
clustering problems linked to, for instance, stock or other financial
data, economic, governmental or medical data as well as machine
monitoring
\citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
The main challenge -- and also the difference with cross-sectional
clustering problems -- is data dimensionality given the multiplicity of
data points for every individual object -- local authority in our case
-- included in the data set. Time-series are dynamic data as the value
of the observations change as a function of time
\citep{aghabozorgi2015time}. This high dimensionality leads to (i)
computational and algorithmic challenges regarding handling these data
and building algorithms to perform clustering over long time-series, and
(ii) open questions regarding the choice of similarity measures in order
to cluster similar times series objects together considering the whole
length of the time-series and overcoming issues around noise, outliers
and shifts \citep{lin2004iterative, aghabozorgi2015time}.

Time-series clustering methods utilising the whole length of time-series
can be grouped in three categories. The first -- model-based approaches
-- is based on recovering the underlying model for each time-series and
then applying clustering algorithms on the model parameters of each
time-series \citep{aghabozorgi2015time}. The main criticism is the
cluster accuracy for nearby clusters \citep{mitsa2010temporal}. The
second approach is a based the formation of vectors of features based on
the original time-series. These new data of reduced dimensionality is
then clustered using conventional clustering algorithms.

For this paper we utilise the third category of time-series clustering
methods known as shape-based approaches. These methods match two sperate
time-series objects based on the similarity of their shapes through the
calculation of distances among these two shapes and are better equipped
to capture similarities between short length time-series
\citep{aghabozorgi2015time}. This approach serves best this paper
because (i) we aim to identify clusters of UK local authorities with
similar temporal signatures -- i.e.~shapes -- of experienced internet
speeds and (ii) the length of our time-series is short (see the data
discussion in this section).

Another import element of time-series clustering is the actual
clustering algorithm. Similar to clustering of cross-sectional data, we
can employ partitioning algorithms, which lead to non-overlapping
clusters, hierarchical clustering, which result to a hierarchy of
clusters and fuzzy algorithms, which create overlapping clusters
\citep{sardatime}. Because of the simplicity of the implementation and
the interpretability of the results, we utilise here partitioning
clustering based on the widely used \emph{k-Means} algorithm. This is an
iterative algorithm, which begins with defining the desired number of
clusters \emph{k}. Then, each observation is randomly assigned to a
cluster from the \([1,k]\) space. This is the initial cluster
assignment, which is followed by iterations in order to minimise the
distance between the centroids of the clusters and the observations
assigned to these clusters \citep{james2013introduction}.

There are a number of differences between the above described
application of \emph{k-Means} for cross-sectional data and its
application for times series data. Instead of creating clusters around
centroids, a common approach is to create clusters around
\emph{medoids}, which are representative time-series objects with a
minimal distance to all other cluster objects \citep{sardatime}. Also,
instead of calculating the Euclidean distance between centroids and data
points, more complex distance measures need to be employed in order to
capture the similarity between a time-series object and a medoid. A
common distance measure for shape-based time-series clustering in the
Dynamic Time Warping (DTW). Using its underpinning dynamic programming
algorithm, DTW compares two time-series objects to find the optimum
warping path between them. DTW is widely used in order to overcome
limitations linked to the use of Euclidean distance
\citep{sardatime, berndt1994using, ratanamahatana2004everything}. The
\texttt{R} package \texttt{dtwclust} has been used for the time-series
clustering \citep{dtwclust}.

To assess the quality and reliability of internet across local
authorities in the UK during the time when the population were told to
work from home if at all possible we utilise unique data regarding
individual internet speed tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
This a private company that allows internet users to check their own
broadband upload and download speeds, and stores every speed-check with
timestamp and geolocation information. These data have been used before
to assess digital divides \citep{riddlesden2014broadband} and the impact
of local loop unbundling regulatory processes
\citep{nardotto2015unbundling}. These volunteered geographic data enable
to assess the \emph{experienced} internet speed by users, which may
differ from the \emph{advertised} maximum speeds provided by Internet
Service Providers. We are particularly interested in upload speeds and
the frequency of speed tests. While the former is less associated with
internet-based leisure activities such as video streaming, which involve
downloading large amounts of data, the latter can be linked more with
work-related activities such as uploading documents or two-way
communication and interaction systems. \textbf{ADD MORE. HOW ABOUT SPEED
TESTS?}

The first step in the workflow after dropping some outliers following
\citet{riddlesden2014broadband} is to transform and the individual,
geolocated and time-stamped tests to more meaningful aggregates both in
terms of space and time. Regarding the temporal dimension, we aggregate
all the speed-checks during the \(13\) weeks of March to May inclusive
for weekdays in \(2020\) by each hour of the day and day of the week. As
our research aims to identify the geography of internet service
resilience for work purposes, bank holidays and the hours between
midnight and 6am were also excluded. The composite week time-series thus
comprise 18 hours multiplied by 5 weekdays or 90 time points per series.
We aggregate these data because, we could not follow individuals or
households and connect data points. Also, although this is a large data
set (\(241,088\) individual tests during the working hours of study
period), there are not enough observations for each local and for each
working hour (\(631\) speed test per LAD in average). These time-series
were calculated for each of the 382 Local Authority Districts (LADs) in
the UK, standardised, and then a \emph{k}-means partitioning around
medoids clustering algorithm was applied using DTW. We initially run the
algorithm for \(k = 5, 10, 15 and 20\) and used cluster validity indices
(CVIs) to pick the optimal solution of \(k = 10\). Following
\citet{sardatime} the following CVIs were employed: Silhouette (max),
Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min),
Modified Davies-Bouldin (DB*, min), Dunn (max), COP (min).

We then match the cluster membership to the LAD-level data to identify
the characteristics of each cluster, including number of LADs, the
descriptive statistics of upload speeds in that cluster, and the
temporal profile by hour of the day and day of the composite week. Thus
we aim to answer the first part of our research question: How resilient
are broadband speeds as experienced in different parts of the UK during
a time of extreme demand?

Since the quality and reliability of internet services vary in time and
space due to both supply and demand-side influences, we use a number of
different measures of experienced upload speeds. These include: a) mean,
experienced connection speed, b) standard deviation or the amount of
fluctuation from the mean, and c) the variation in speeds at particular
times of day when working from home is more likely to take place. We
take account of all three measurements in order to describe upload
speeds as fully as possible. The following \textbf{graphs / tables} show
these statistics for the sample as a whole, as well as similar
statistics related to frequency of testing.

@@

\textbf{Just looking back, it was partly the frequency of tests, but
that was also by geography, so do we need maps? And I feel there was
something else I'm forgetting\ldots{}}

The cause of these different experiences of broadband resilience may be
different in different areas, as they may reflect either similarities in
patterns of demand or similar quality of infrastructure. Our approach is
also limited by potential endogeneity, as for example, better quality
connections with high mean speeds may enable more working from home, but
greater demand may cause slower speeds, less reliability and greater
variability of speed at different times of day or week. Therefore, we
avoid attributing any cause to our analysis of the experienced level of
quality and reliability of upload speeds. Instead, we run an auxiliary
regression in order to understand how the spatial and temporal patterns
of internet service relate to the economic geography of the UK. We
discuss how the different patterns might support or undermine efforts to
work from home and maintain safe productivity and whether they reinforce
existing spatial and social inequalities. More specifically, we estimate
the following multinomial model:

\begin{align}
Pr(Y_{i}=j) = \frac{exp^{X_{i}\beta_{j}}}{\sum_{i=1}^j exp^{X_{i}\beta_{j}}}
\begin{cases}
    i = 1, 2, ... , N \\t
    j = 1, 2, ... , J
\end{cases}\label{eq1}
\end{align}

The identification strategy is as follows. Based on the outcomes of the
time-series clustering, we identify \(J\) distinct and crisp clusters.
We then regress this cluster membership against a vector \(X_{i}\) of
socio-economic and geographic variables, which we are discussed in
details in \textbf{section XX}. This analysis enables us to provide a
more nuanced understanding of how telecommuting and technology intersect
at a time of extreme demand, and what lessons this time has for a future
where telecommuting is likely to remain a common means of accessing work
and broadband services, as well as infrastructure, must be fit for
purpose.

@@

The temporal profiles used to cluster the local authorities have been
summarised in graph {[}.{]}, which shows a composite profile of mean
upload speeds per hour per day for each cluster. For upload speeds, 345
of 382 local authorities, or over 62 million people, fall into cluster 6
or cluster 9. Graph {[}.{]} suggests that these two clusters have
relatively similar temporal profiles, which are flatter than the other,
smaller clusters, suggesting better reliability of service. However, the
upload speeds at all times for cluster 9 are substantially higher than
for cluster 6, which is an indication of better quality of service. This
difference is reflected not only in the mean speeds for these clusters
for the whole sample, but also the mean upload speeds in the morning
peak from 9:00-10:59, as well as the evening peak period from
19:00-20:59. In comparison, the time profile in graph {[}.{]} shows
upload speeds in cluster 1 are on average lower at certain times of day
during the study period than any other cluster, whilst the profile for
cluster 3 appears to show speeds fluctuating as much as cluster 1, but
at levels usually higher even than cluster 9.

The variability of the smaller clusters may be related to the fewer
speed tests from fewer local authorities that have been averaged, whilst
averaging greater numbers of speed tests could artificially flatten the
profile. This appears not to be the case for cluster 9, as there is
negligible difference between morning and evening upload speeds, at 1\%
slower in the morning, confirming a high level of reliability of
service. In comparison, upload speeds in cluster 6 are 4\% slower in the
morning than in the evening and experienced the joint highest ratio of
standard deviation to mean across the time period under assessment. This
suggests that although the time profile is relatively flat in graph
{[}.{]}, the experience is one of speeds that fluctuate from a lower
mean, and therefore might more often impact on online activities. Still,
in most of the smaller clusters the reliability of service is worse than
in cluster 9 in terms of the ratio of standard deviation to mean. The
smaller clusters are mainly worse than clusters 9 and 6 in terms of the
ratio of upload speed in the morning peak compared to the evening peak.
In five of the smaller clusters, which are home to almost 3 million
people, speeds are 12-19\% slower between 9:00-10:59 compared to
19:00-20:59.
